---
title: "Group 5 ASSG2"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,error = FALSE, message = FALSE,
                      warning = FALSE,
                      out.height = "\\textheight",  out.widtibh = "\\textwidth")
```

```{r packages, include=FALSE}
library(dplyr)
library(psych) # For describe() providing summary stats
library(quantmod) # For importing stock from Yahoo
library(ggplot2) # Visualizations
library(forecast) # Autoplot
library(tseries) # ADF test and more
library(FinTS) # ARCH LM test
```

**Importing Ford Motor Company stock statistics**

```{r importing, echo=TRUE}
#This code does it.
getSymbols('F', src = 'yahoo', from = as.Date('2022-01-01'), to = 
             as.Date('2025-03-20'))
```
**Some Ford's stock closing prices**

```{r closing prices, echo=FALSE}
F_closing <- F$F.Close
head(F_closing)
```

**Time series plot of Ford's Closing prices**

```{r ts conversion, echo=FALSE}
F_closing_ts <- ts(F_closing, start = c(2022,1), frequency = 240)
autoplot(F_closing_ts)
```


```{r daily log returns, echo=TRUE}
#This computes the daily log returns
F_returns <- diff(log(F_closing_ts))
```

**PART 1: EXPLORATORY DATA ANALYSIS**

**Log-returns ts plot**

```{r returns plot, echo=FALSE}
autoplot(F_returns) + 
  ggtitle('Ford Motor Stock daily log returns')
```

**Summary statistics of the returns**

```{r summary statistics, echo=FALSE}
describe(F_returns)
```

*Ford Company's stock returns have a mean of zero and a variance of 1. This can imply stationarity since these statistics do not change over time, but further tests are needed to truly confirm this.*

-Skewness measures asymmetry. This tells us whether there are extreme values on the left or on the right.

*The negative skewness of -1 suggests the returns have a slightly longer left tail. Large losses are therefore more likely to occur.*

-Kurtosis measures the tailedness of a distribution. This shows how often extreme values occur as compared to a normal distribution (Where kurtosis = 3).

*Results show that returns have a high kurtosis (returns are leptokurtic). Losses (or gains) occur more frequently than normal. This implies higher risk involved with this stock.*

**ADF Stationarity test**

-Null hypothesis: The Series is non-stationary

*Fail to reject if P > [level of significance]*

-Alternative hypothesis: The Series is stationary

*Reject Null in favour of the alternative if P < [level of significance]*

-The results are as shown:

```{r adf, echo=FALSE}
print(adf.test(F_returns))
```

-Assuming the default significance level of 5% (0.05), The p-value shown is less than this. We therefore reject the null hypothesis in favour of the alternative one and conclude that Ford's returns are indeed stationary.

**Checking for ARCH effects on squared returns**

```{r ARCH effects, echo=FALSE}
ggAcf(F_returns^2) + 
  ggtitle('Autocorrelations of Ford Company squared returns')
```

-Only 7 out of 480 lags exceed the confidence interval(~1.458%). At 5% significance, 24 lags (480 * 5%) or more exceeding the confidence level would be regarded as statistically significant. We can therefore safely ignore the lags appearing above the ci as there are statistically insignificant

-The above plot therefore shows no significant autocorrelations seen from the squared returns; and therefore no volatility clustering. This means that volatility (squared returns) is homoscedastic (not heteroscedastic)... the variance, or volatility, is therefore constant.

-ARCH effects can further be tested in depth using the Lagrange Multiplier (LM) test. 

**ARCH-LM test**

-Null hypothesis: No ARCH effects (homoscedasticity)

-Alternative hypothesis: There is ARCH effects (heteroscedasticity)

*Reject Null hypothesis if P < [level of significance]*

```{r ARCH-LM, echo=FALSE}
print(ArchTest(F_returns))
```
-The p-value (0.9881) is significantly more than 0.05. We therefore fail to reject the Null hypothesis and conclude that Ford Company's stock returns are homoscedastic; they exhibit no ARCH effects.

**PART 2 ARCH**

**Estimating ARCH(p)**


